**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-02-03 Manuscripts/Manuscript_AM_v2.Rmd
updated manuscript and references

When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004]. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study. Another less known side effect of multiple comparisons is that they reduce statistical power ($\alpha$)[@nakagawa2004], which in turn can lead to fail to reject the null hypothesis although it is false (a type II error)[@albers2019].

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population. When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].

One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006]. The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend. In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation. In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time. 
Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data. In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior. In such scenario, the _p-value_ becomes invalid[@wang2019].

As  the “frequentist” rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.

The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures. Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.  Third, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario. Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.  In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.

>>>>

When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004]. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate (Type I error) [@liu2010;@halsey2015], consequently biasing the conclusions of the study. Although corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), they in turn reduce statistical power ($\alpha$)[@nakagawa2004], and lead to increased Type II error (fail to reject the null hypothesis although it is false [@gelman2012;@albers2019]. Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I and II errors.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].

One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006]. The linearity assumption in both rm-ANOVA and LMEMs  implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit LMEMs or rm-ANOVA model inconsistent with the observed variation. Therefore, when LMEMs and rm-ANOVA are used to draw inference of such highly-variable data the estimates are inevitably biased, because non-linear data is provided to a model which can only accomodate linear trends.

Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) the validity of such metric relies on how adequate the model fits the data. In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019]. For example, a model that is only able to explain linear responses but is fitted to data that has a quadratic behavior results in _p-values_ and parameter estimates that are invalid[@wang2019].

As  the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed behavior) to determine the trend in the data, which enables inference that is appropriate.

In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations, and can model non-constant correlation between repeated measurements [@wood2017]. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding missing observations, assumption of linearity of response and constant correlation structures is explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, simulated data that reproduces patterns in previously reported studies [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that often occurs in biomedical research. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.
